The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-04 pkg/partykit/inst/guideliketest/draft/draft.Rnw
fine-tuning of layout and pagebreaks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2607 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7662f70e5bc36c48644010a8983fb7e82dda480a

\begin{table}
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_{k-1}(Z_1,Z_2)$, $k=1,2=K$.}
\end{table}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=3>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=3>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
@
\endminipage

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=3>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

\subsubsection{``Tree'' scenario}
For the second simulation scenario (``tree'' scenario) the same basic setting with the same 
variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
on $[-1,1]$. Therefore, an additional split
is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
to a change in the intercept $\beta_0$.
Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
%
\begin{align*}
\beta_{0}(Z_1, Z_2) = \begin{cases}
0       \qquad \text{if } Z_2 \leq \xi \\
-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
\end{cases}
\end{align*}
\begin{align*}
\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
+\delta \quad \text{if } Z_2 \leq \xi \\
-\delta \quad \text{if } Z_2 > \xi
\end{cases}
\end{align*}
%
The underlying tree structure for the data generating process is illustrated in 
Figure~\ref{fig:dgp_tree}.

<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=

# d <- dgp_tree(nobs = 250)
set.seed(7)
d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)

@@

\begin{figure}%[t]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5, height=4.5>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure
applied in the data generating process for the ``tree'' scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
\hline
Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without dichotomization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
                             &                     & \quad indicating the potential split point\\
                             & \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l l l}
\hline\noalign{\smallskip} 
Residuals/Scores   & Dichotomization & Categorization & \multicolumn{2}{l}{Via testing strategy}\\
\noalign{\smallskip}\hline\noalign{\smallskip}
residuals	   & yes	     & cat            & GUIDE & {(default)}\\
residuals	   & yes	     & max            & MOB   & {(with modifications)}\\
residuals	   & yes	     & lin            & CTree & {(with modifications)}\\
residuals	   & no 	     & cat            & MOB   & {(with modifications)}\\
residuals	   & no 	     & max            & MOB   & {(with modifications)}\\
residuals	   & no 	     & lin            & CTree & {(with modifications)}\\
scores  	   & yes	     & cat            & GUIDE & {(with modifications)}\\
scores  	   & yes	     & max            & MOB   & {(with modifications)}\\
scores  	   & yes	     & lin            & CTree & {(with modifications)}\\
scores  	   & no 	     & cat            & MOB   & {(with modifications)}\\
scores  	   & no 	     & max            & MOB   & {(default)}\\
scores  	   & no 	     & lin            & CTree & {(default)}\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Testing strategies applied for all 12 factorial combinations of the building blocks.
For each combination one of the three testing strategies CTree/MOB/GUIDE is applied,
possibly after modifying the assessed variables as indicated (i.e., dichotomized or categorized).
\label{tab:combinations_appendix}}
\end{table}

@@

\begin{figure}[t]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree", "MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

@@

\begin{figure}%%[t]
\setkeys{Gin}{width=\linewidth}
<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
#subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
                       labels = c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores"))
par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                             fill = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]), 
                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]))

@@

\begin{figure}%[t]
\setkeys{Gin}{width=\linewidth}
<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
                                         #"ctree_max", "ctree_max_bin",
                                         "mfluc", "mfluc_bin",
                                         "guide_sum_12"))

@@

@
\caption{\label{fig:bin_stump}
Selection probability of the true split variable $Z_1$ for
testing strategies CTree, CTree+dich, MOB, MOB+dich, and GUIDE+scores in the ``stump''
scenario with both parameters varying. CTree+dich and MOB+dich refer to modifications
of CTree and MOB applying a dichotomization of residuals/scores.
Probabilities are estimated over increasing effect size $\delta$
with 100 replications of 250 observations per step. 
Each panel represents a different value of the true split point $\xi$ (0, 0.2, 0.5 
and 0.8 from left to right corresponding to the 50\%, 60\%, 75\%, and 90\% quantile).
}
\end{figure}

>>>>

\begin{table}[t!]
\centering
\begin{tabular}{l l l}
\hline
Name                    & Notation                   & Specification \\
\hline
\textit{Variables:}     &                            & \\ 
Response                & $Y$                        & $=\beta_0(Z_1) + \beta_1(Z_1) \cdot X + \epsilon$ \\
Regressor               & $X$                        & $\mathcal{U}([-1,1])$ \\
Error                   & $\epsilon$                 & $\mathcal{N}(0,1)$ \\
True split variable     & $Z_1$                      & $\mathcal{U}([-1,1])$\\
Noise split variables   & $Z_2, Z_3, \dots, Z_{10}$  & $\mathcal{U}([-1,1])$  or $\mathcal{N}(0, 1)$ (alternating) \\
\hline
\textit{Parameters/functions:} &            & \\ 
Intercept                      & $\beta_0$  & $0$ or $\pm \delta$\\
Slope                          & $\beta_1$  & $1$ or $\pm \delta$\\
True split point               & $\xi$      & $\in \{0, 0.2, 0.5, 0.8\}$ \\
Effect size                    & $\delta$   & $\in \{0, 0.1, 0.2, \ldots , 1\}$ \\
\hline
\end{tabular}
\caption{\label{tab:variables}Variables included in the data generating process 
as used for the ``stump'' scenario. In the ``tree'' scenario $Z_2$ is also a true split variable, 
not only $Z_1$, and hence $\beta_0$ and $\beta_1$ are functions $\beta_{k-1}(Z_1,Z_2)$, $k=1,2=K$.}
\end{table}

@@

\begin{figure}[p!]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep, terminal_panel = tp)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 3.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = 2.2)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.2)
mtext(expression(beta[1]== 1), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = -0.8)
@
\endminipage

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
# d <- dgp_tree(nobs = 250)
set.seed(7)
d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)

@@

\begin{figure}[p!]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5.2, height=4.4>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure
applied in the data generating process for the ``tree'' scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

\subsubsection{``Tree'' scenario}

For the second simulation scenario (``tree'' scenario, see Figure~\ref{fig:dgp_tree}) the same basic setting with the same 
variables as in the ``stump'' scenario is applied (see Table~\ref{tab:variables}).
However, not only $Z_1$ but also $Z_2$ is used as a true split variable following a uniform distribution
on $[-1,1]$. Therefore, an additional split
is preformed yielding a tree with three terminal nodes. The first split (in $Z_2$, at~$\xi$) induces a 
change in the slope parameter $\beta_1$ while the second split (in $Z_1$, also at~$\xi$) corresponds 
to a change in the intercept $\beta_0$.
Hence, contrary to the ``stump'' scenario where three variations are considered for the parameters 
$\beta_0$ and $\beta_1$, only this one variation with both parameters varying is investigated for 
the ``tree'' scenario. In particular, the parameters depend on the split variables in the following way:
%
\begin{align*}
\beta_{0}(Z_1, Z_2) = \begin{cases}
0       \qquad \text{if } Z_2 \leq \xi \\
-\delta  \quad \text{ if } Z_1 \leq \xi \land Z_2 >\xi \\
+\delta  \quad \text{ if } Z_1 > \xi \land Z_2 >\xi
\end{cases}
\end{align*}
\begin{align*}
\beta_{1}(Z_1, Z_2) = \beta_{1}(Z_2)= \begin{cases}
+\delta \quad \text{if } Z_2 \leq \xi \\
-\delta \quad \text{if } Z_2 > \xi
\end{cases}
\end{align*}

@@

\begin{table}[t!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
\hline
Dichotomization              & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without dichotomization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)^{\top}$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)^{\top}$\\
                             &                     & \quad indicating the potential split point\\
                             & \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\begin{table}[p!]
\centering
\begin{tabular}{ l l l l l}
\hline\noalign{\smallskip} 
Residuals/Scores   & Dichotomization & Categorization & \multicolumn{2}{l}{Via testing strategy}\\
\noalign{\smallskip}\hline\noalign{\smallskip}
residuals	   & yes	     & cat            & GUIDE & {(default)}\\
residuals	   & yes	     & max            & MOB   & {(with modifications)}\\
residuals	   & yes	     & lin            & CTree & {(with modifications)}\\
residuals	   & no 	     & cat            & MOB   & {(with modifications)}\\
residuals	   & no 	     & max            & MOB   & {(with modifications)}\\
residuals	   & no 	     & lin            & CTree & {(with modifications)}\\
scores  	   & yes	     & cat            & GUIDE & {(with modifications)}\\
scores  	   & yes	     & max            & MOB   & {(with modifications)}\\
scores  	   & yes	     & lin            & CTree & {(with modifications)}\\
scores  	   & no 	     & cat            & MOB   & {(with modifications)}\\
scores  	   & no 	     & max            & MOB   & {(default)}\\
scores  	   & no 	     & lin            & CTree & {(default)}\\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Testing strategies applied for all 12 factorial combinations of the building blocks.
For each combination one of the three testing strategies CTree/MOB/GUIDE is applied,
possibly after modifying the assessed variables as indicated (i.e., dichotomized or categorized).
\label{tab:combinations_appendix}}
\end{table}

@@

\begin{figure}[p!]
\setkeys{Gin}{width=\linewidth}
<<scores_stump_wosign, fig=TRUE, echo=FALSE, height=5, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "mfluc", "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree", "MOB",
                                  "GUIDE+scores", "GUIDE"))

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
#subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
                       labels = c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores"))
par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                             fill = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]), 
                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]))

@@

@@

\begin{figure}[h!]
\setkeys{Gin}{width=\linewidth}
<<bin_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_bin", 
                                         #"ctree_max", "ctree_max_bin",
                                         "mfluc", "mfluc_bin",
                                         "guide_sum_12"))

@@

@
\caption{\label{fig:bin_stump}
Selection probability of the true split variable $Z_1$ for
testing strategies CTree, CTree+dich, MOB, MOB+dich, and GUIDE+scores in the ``stump''
scenario with both parameters varying. CTree+dich and MOB+dich refer to modifications
of CTree and MOB applying a dichotomization of residuals/scores.
Probabilities are estimated over increasing effect size $\delta$
with 100 replications of 250 observations per step. 
Each panel represents a different value of the true split point $\xi$ (0, 0.2, 0.5 
and 0.8 from left to right corresponding to the 50\%, 60\%, 75\%, and 90\% quantile).
}

\setkeys{Gin}{width=\linewidth}
<<cat_stump, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_cat",
                                         #"ctree_max", "ctree_max_cat",
                                         "mfluc", "mfluc_cat",
                                         "guide_sum_12"))
subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_cat", 
                                  #"ctree_max","ctree_max_cat",
                                  "mfluc", "mfluc_cat",
                                  "guide_sum_12"),
                       labels = c("CTree","CTree+cat",
                                  #"CTree+max","CTree+max_cat",
                                  "MOB", "MOB+cat",
                                  "GUIDE+scores"))
